---
title: "Summarising Your Data"
author: "Andrew Stewart, Email: drandrewjstewart@gmail.com, Twitter: @ajstewart_lang"
date: ""
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

Once a dataset has been tidied, often one of the first things we want to do is generate summary statistics. We'll be using the `mpg` dataset that is built into the `tidyverse` for this workshop. This dataset contains information about cars (such as engine size, fuel economy) produced by a number of different manufacturers. How would be go about generating (e.g.) the means and standard deviations grouped by car manufacturer for one of our variables? Have a look at the following video where I walk you through this worksheet. Then I want you to work through the content by writing (and running) the script on your own machine.

&nbsp;&nbsp;

<center>

<iframe width="560" height="315" src="https://youtube.com/embed/8P3E-lY15a8" frameborder="0" allowfullscreen></iframe>

</center>

&nbsp;&nbsp;

Remember to set up a .Rproj file for this workshop before continuing. In your script, you'll first need to load the tidyverse. 

```{r, message = FALSE}
library(tidyverse)
```

# Using `group_by()` and `summarise()`

We are going to use the `group_by()` function to group the dataset, and then the `summarise()` function to calculate the mean of hte `hwy` variable. The `summarise()` function can take a lot of different functions to give us summary statistics. To read more about the different options, type `?summarise` in the Console window. Commonly used ones are `mean()`, `median()`, `sd()`.

```{r}
mpg %>%
  group_by(manufacturer) %>%
  summarise(mean_hwy = mean(hwy), sd_hwy = sd(hwy), number = n())
```
Note that this output is currently ordered alphabetically by the first column `manufacturer`. What if we wanted to order this out by mean highway fuel economy highest (best) to lowest (worst)? We can use the arrange function.

## Re-ordering the output with `arrange()`

```{r}
mpg %>%
  group_by(manufacturer) %>%
  summarise(mean_hwy = mean(hwy), sd_hwy = sd(hwy), number = n()) %>%
  arrange(mean_hwy)
```

Hmm, so that isn't what we want - this is going from lowest to highest which is the default in R. We can change that by putting a `-` sign in from of the parameter we can to order by.

```{r}
mpg %>%
  group_by(manufacturer) %>%
  summarise(mean_hwy = mean(hwy), sd_hwy = sd(hwy), number = n()) %>%
  arrange(- mean_hwy)
```

This is looking better.

## The `summarise_at()` variant

As well as using `summarise()`, you can use related functions such as `summarise_at()`. This is a scoped version of the `summarise()` function that can be applied across multiple columns. Note, when using `summarise_at()` you need to put the columns you want to summarise over in quotes. You also need to provide the summary function - in this case `mean`. Finally, in case our dataset contains any missing values (indicated by `NA`), we set the parameter `na.rm = TRUE`. This will ensure that missing data points are removed before the operation is applied. If we had missing data, but didn't tell R what we wanted to do with it, it would have thrown an error. 

```{r}
mpg %>% 
  group_by(manufacturer) %>%
  summarise_at(c("displ", "cty", "hwy"), mean, na.rm = TRUE)
```

## The `summarise_if()` variant

Imagine we had a really big dataset and wanted to summarise all columns that were of a certain type. We can use the `summarise_if()` function to work out the mean for each of our car manufactures as follows:

```{r}
mpg %>% 
  group_by(manufacturer) %>%
  summarise_if(is.numeric, mean, na.rm = TRUE)
```

The first parameter in `summarise_if()` is the logical test applied to each column - in this case, if a column is numeric (i.e., a number) - then the test evaluates to TRUE and the second function, `mean`, is applied. Again, we tell R to ignore missing (`NA`) data values with the `na.rm = TRUE` parameter. 

R functions differ in terms of what arguments they take. I often forget them - if you start typing a function name, you'll get a little bubble above where you're typing to remind you what parameters are needed. And if you can't remember the details, you can just type `help(function_name)` or `?function_name` in the console for any function that you need help with. A lot of data analysis with R (or Python or wny other language really) involves a fair bit of Googling. This is normal. There are some things I can never remember and am always having to look up!

# Adding columns using `mutate()`

We can add a new column that I'm calling `mean_hwy` using the `mutate()` function like this.

```{r}
mpg %>% 
  group_by(manufacturer) %>%
  mutate(mean_hwy = mean(hwy), sd_hwy = sd(hwy))
```

We have too many columns to display on this page so we can drop a coouple by using the `select()` function slightly differently. By putting a `-` sign in front of a column names in `select()` we end up dropping it.

```{r}
mpg %>% 
  group_by(manufacturer) %>%
    mutate(mean_hwy = mean(hwy), sd_hwy = sd(hwy)) %>%
  select(-class, -trans)
```

Note that this doesn't change the `mpg` dataset permanently - the changes won't be saved unless we map the output of this code onto a new variable. Below I am doing this by using the assignment operator `<-` to map it onto a new variable I'm calling `mpg_with_mean`. Note that we remove the grouping at the end as we don't want our grouping rule to remain in our new data frame.

```{r}
mpg_with_mean <- mpg %>% 
  group_by(manufacturer) %>%
    mutate(mean_hwy = mean(hwy), sd_hyw = sd(hwy)) %>%
  ungroup() %>%
  select(-class) 
```

We can then inspect this new variable using `head()` and `str()`.

```{r}
head(mpg_with_mean)
```

```{r}
str(mpg_with_mean)
```

# Your challenge

The `Tidyverse` has a number of other built-in data sets. Another one is the `starwars` data set. You can have a look at it by typing `starwars` or by typing `view(starwars)`. This second option will open the data set in a new window. Have a go playing around with it. Work out the mean height of humans in the Star Wars universe. There might be some missing data (indicated by `NA`). You can use the `na.rm = TRUE` parameter in your `summarise()` function to ignore these values when generating your summary statistics. 

Another way to filter out `NA` values is to use the `filter()` function in your pipeline. The function `is.na()` returns a logical value of TRUE of FALSE. The operator `!` means NOT, so the expression `!is.na(height)` will return TRUE when the height value is present, and FALSE if absent. By combining this with `filter()` we have the line `filter(!is.na(height))` which will filter only the cases where we have height data (i.e., `!is.na(height)` is TRUE). So your code might look like this:

```{r, eval=FALSE}
starwars %>%
  filter(!is.na(height)) %>%
  filter(species == "Human") %>%
  summarise(mean_height = mean(height))
```

Replace the word `mean` in the `summarise()` line with `median`. What other things can you replace it with? Hint: type `?summarise` in the console.  What other summary information can you extract from this dataset?

## Improve this Workshop

If you spot any issues/errors in this workshop, you can raise an issue or create a pull request for [this repo](https://github.com/ajstewartlang/04_summarising_your_data). 
